Within chert: a multi-technique mineral and geochemical approach to the study of chert of southwestern Iberia

### 2025-04-26 analysis/paper/paper.qmd
Improvements to intro

In the Iberian Peninsula - a key region for studying human adaptations and cultural transitions - these approaches have also been applied to UP archaeological assemblages. These multi-method strategies have successfully characterized cherts from diverse geological formations and identified mobility patterns throughout the Upper Paleolithic, including evidence of long-distance contacts between hunter-gatherer groups.

In recent years, a wide range of analytical techniques - such as X-Ray Diffraction (XRD), X-Ray Fluorescence (XRF), portable XRF, Laser Ablation Inductively Coupled Plasma Mass Spectrometry (LA-ICP-MS), Raman Spectroscopy, and Scanning Electron Microscopy with Energy Dispersive Spectroscopy (SEM-EDS) - has been employed to enhance chert characterisation in this region. Each method offers distinct advantages in data resolution, sample preparation, and cost-effectiveness. However, no single technique is universally applicable, and a multi-analytical approach is often necessary to effectively address specific research questions.

Southern Portugal, while having benefited from previous geochemical studies of one of the four chert-bearing formations, as well as macroscopic and petrographic studies of chert-bearing formations (Belmiro et al., 2023) and Upper Palaeolithic chert assemblages from the site of Vale Boi (Belmiro et al., 2025), remains largely unexplored through other analytical techniques. These may be key to address open research topics and questions by recent studies.

For example, local and regional Jurassic chert sources are abundant in the region and in relative proximity to the site, yet the differentiation between specific chert formations remains unclear, which is critical for understanding regional mobility patterns (Belmiro et al., 2023).

Additionally, evidence suggests that non-local materials were frequently introduced into the archaeological site, indicating complex long-distance procurement dynamics. However, these findings were based on macroscopic and petrographic studies, which leaves room for further characterization to strengthen source attributions and test current hypotheses.

Following the results and questions from previous studies by our team (Belmiro et al., 2023; Belmiro et al., 2025), this study employs an integrated analytical approach to geological chert samples from local, regional, and non-local formations, as well as archaeological assemblages from the UP levels of Vale Boi to accomplish three main goals:

@@

By refining the characterization of chert in the region and improving source attribution at the site, this research contributes to broader discussions on lithic procurement, human mobility, and exchange networks in southwestern Iberia during the UP.

@@

For this experiment, a Bruker portable XRF Titan S1 was used in a laboratory benchtop setup, powered by battery (with up to 25% battery charge before being replaced by a fully charged battery). The standard databases available in the equipment from Bruker were employed: Geochem application and the Dual Mining method. A detailed description of the preparation and analysis process is available in our online repository (SOM_1).

Chert samples from various sources and types, including both geological and archaeological specimens, were scanned (Table 3). After scanning, the thickness and diameter of the scanned face were measured to ensure that each sample met the minimum required thickness and size (Newlander et al., 2015; SOM ++).

@@

The results presented in this section focus on the main observed patterns of the analysis between outcrops, formations and chert types. The detailed analysis/graphs regarding individual samples for all methods can be found in our online repository (SOM_2).

@@

The pXRF measured 44 major and minor elements, although, for a large majority, these were mostly composed of values below the detection level (\< LOD) or with values equal to 0. Only elements with more than 50% (n\>80) of successful values were maintained in the analysis to allow the successful interpretation of the data (Fig. ++). This list and a summarised description of the elements and their presence in sedimentary rocks/chert can be found in the supplementary materials (SOM ++).

>>>>

As seen in the previous chapters, we have applied macroscopic and petrographic approaches to the study of cherts from the Algarve region (Belmiro et al., 2023, 2025). However, the chert-bearing formations from the Algarve region remain largely unexplored through other analytical techniques.

An exception would be the Lower Jurassic (Carixian) and Upper Jurassic (Malm) limestone and dolomitic limestone formations from western Algarve [@rochaNoticiaExplicativaFolha1979], which show previous mineral and geochemical characterisations [@ribeiro_evolucao_2005]. However, given the geological scope of the study, only 3 chert samples from the Carixian and 3 samples of the Malm were analysed. While providing a basis for mineral and chemical composition of these cherts, it is key to expand the analysis to a more robust sample size.

Similarly, the Upper Palaeolithic chert assemblages from the site of Vale Boi have yet to be further characterised through geochemical methods. These may be key to address open research topics and questions by recent studies focusing on chert provisioning and selection during the UP at Vale Boi.

For example, the Lower Jurassic (Carixian) cherts, located near Vale Boi (\<20 km) show macroscopic and petrographic similarities to the Middle Jurassic (Malhão formation) cherts located eastwards (\>70 km). This makes the differentiation between specific chert formations unclear. Making this distinction is critical for understanding regional mobility patterns and whether local cherts at Vale Boi are a mix of local (\<30 km) and regional (xkm) resources (Belmiro et al., 2023).

Questions regarding nodule selection may also be further explored to further characterise close-range chert procurement. This is particularly relevant for the Lower Jurassic (Carixian) cherts which show different degrees of dolomitization, with a clear impact on the knapping quality. By further characterising the mineral and chemical composition of geological cherts from all outcrops and macroscopic variability, and comparing with the archaeological samples, we may explore behaviours associated to selection associated with nodule quality (sensu higher silicification).

Additionally, evidence suggests that non-local materials were frequently introduced into the archaeological site, indicating complex long-distance procurement dynamics. However, these findings were based on macroscopic and petrographic studies, which leaves room for further characterization to strengthen source attributions and test current hypotheses. This is especially relevant for archaeological chert types for which we did not have a direct comparison regarding petrography, as for example T7 possibly originating from the Betic systems (south Spain) or other types for which a source could not be identified (T8-11).

To further explore these questions and topics, this study employs an integrated analytical approach to geological chert samples from local, regional, and non-local formations, as well as archaeological assemblages from the UP levels of Vale Boi to accomplish the following goals:

@@

By refining the characterization of chert in the region, improving source attribution at the site, and identifying the strengths and weaknesses of different analytical methods, this research contributes to broader discussions on lithic procurement, human mobility, and exchange networks in southwestern Iberia during the UP.

@@

For this experiment, a Bruker portable XRF Titan S1 was used in a laboratory benchtop setup, powered by battery (with up to 25% battery charge before being replaced by a fully charged battery). The standard databases available in the equipment from Bruker were employed: Geochem application and the Dual Mining method. A detailed description of the preparation and analysis process is available in our online repository (Online resource 1).

Chert samples from various sources and types, including both geological and archaeological specimens, were scanned (Table 3). After scanning, the thickness and diameter of the scanned face were measured to ensure that each sample met the minimum required thickness and size (Newlander et al., 2015; Online resource 2).

@@

The results presented in this section focus on the main observed patterns of the analysis between outcrops, formations and chert types. The detailed analysis/graphs regarding individual samples for all methods can be found in our online repository (Online resource 4).

@@

The pXRF measured 44 major and minor elements, although, for a large majority, these were mostly composed of values below the detection level (\< LOD) or with values equal to 0. Only elements with more than 50% (n\>80) of successful values were maintained in the analysis to allow the successful interpretation of the data (Fig. ++). This list and a summarised description of the elements and their presence in sedimentary rocks/chert can be found in the supplementary materials (Online resource 3).
